---
title: Mechanisms through which exercise reduces symptom severity and/or functional impairment in posttraumatic
  stress disorder (PTSD)
author: "Simonne Wright, Virginia Chiocchia, Georgia Salanti <span style='font-size: 15px'></style>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
 html_document:
   df_print: paged
   mainfont: Arialh
---

```{r  setup, error=FALSE, warning=FALSE, message=FALSE, include=FALSE}

#Load package 'meta' to run meta-analyses.
library(meta)
library(tidyr)
library(tibble)
library(stringr)
library(readxl)
library(grid)
library(dplyr)
library(Matrix)
library(knitr)
library(kableExtra)
library(rms) 
library(meta)
library(devtools)
library(tools)
library(metafor)
library(gtools)
library(metasens)
library(robvis)

#setwd("C:/Users/simon/Desktop/Github rep/LSR2_exercise_H/")

load("data/data_2023-12-4_LSR2_Awks2.RData")
load("data/LSR2weeksStata2insufdata.RData")

load("data/LSR2_studychartab_wks_140124.RData")
load("data/LSR2_studychartab_insuf_140124.RData")

load("data/LSR2_interventions_150224.RData")

## prisma<-read.csv("data/LSR3_H_PRISMA_2024.01.23.csv") #csv file for the prisma2020 -->
## prisma<-PRISMA_data(prisma) 

#risk of bias data
rob <- read_xlsx("data/LSR2_robptsd_260124.xlsx")

#Round off results to two digits
settings.meta(digits = 2)

#add ROB variable "high" (no = low/some concerns; yes = high)
data_2023_12_4_LSR2_Awks2$high <- c("no", "yes" , "no" , "yes" , "no" , "yes" , "yes", "yes" , "yes")

#create subset for ROB sensitivity analysis: select studies with treatments "OLA" and "PLA"
data.rob.sens = subset(data_2023_12_4_LSR2_Awks2, high == "no")

```

# Results

## Flow of Study Selection and Descriptives

We identified *number of records* records. See Figure 1 for the PRISMA flow diagram. A total of `r length(table(data_2023_12_4_LSR2_Awks2$studyN))+length(table(LSR2weeksStata2insufdata$studyN))` studies with data from `r sum(data_2023_12_4_LSR2_Awks2$totNrandom, na.rm=T)+ sum(LSR2weeksStata2insufdata$totNrandom, na.rm=T)` participants were eligible for inclusion. 

*prisma diagram*

**Figure  1:** PRISMA flow diagram 

10 of the studies were RCTs, and 1 was a crossover RCT (Greene & Petruzzello, 2022). `r length(table(data_2023_12_4_LSR2_Awks2$studyN))` of the `r  length(table(data_2023_12_4_LSR2_Awks2$studyN)) + length(table(LSR2weeksStata2insufdata$studyN))` studies were included in the meta-analyses (Bryant et al., 2023; Huseth, 2021; Nordbrandt et al., 2020; Rosenbaum et al., 2015; Voorendonk et al., 2023; Whitworth et al., 2019a; Whitworth et al., 2019b; Young-McCaughan et al., 2022). Meta-analysis was not feasible for `r length(table(LSR2weeksStata2insufdata$studyN))` of the `r length(table(data_2023_12_4_LSR2_Awks2$studyN))+length(table(LSR2weeksStata2insufdata$studyN))` studies (Crombie et al., 2021a; Greene & Petruzzello, 2022; Powers et al., 2015). These were synthesized descriptively without meta-analysis. Three of the studies provided follow-up PTSD outcome data which were insufficient for synthesis (Bryant et al., 2023; Voorendonk et al., 2023; Young-McCaughan et al., 2022).Three studies examined putative mediators which was not sufficient to carry out a meta-analysis (Crombie et al., 2021a; Powers et al., 2015; Whitworth et al., 2019b).

Study characteristics of the `r length(table(data_2023_12_4_LSR2_Awks2$studyN))` studies included in the meta-analysis are summarized in Table 1. From these `r length(table(data_2023_12_4_LSR2_Awks2$studyN))` studies, there were `r length(table(data_2023_12_4_LSR2_Awks2$study))` eligible comparisons. One study presented findings from two independent comparisons, including a total of four distinct intervention groups (Young-McCaughan et al., 2022). One of the first authors published findings from two different but methodologically similar trials in the same year (Whitworth et al., 2019a; Whitworth et al., 2019b).

```{r echo=FALSE}
knitr::kable(LSR2_studychartab_wks_140124[, 1:13], "pipe")
```

*Aerobic exercise = physical performance behaviour pattern that increases heart rate and respiration while using large muscle groups repetitively and rhythmically; anaerobic exercise = physical performance behaviour pattern that is performed in short intense bursts with limited oxygen intake; mixed exercise = combination of aerobic and anaerobic exercise. CAPS-IV = Clinician-Administered PTSD Scale - 4th edition; PCL-4 = PTSD Checklist - version 4; PCL-5 = PTSD Checklist version 5; HTQ = Harvard Trauma Questionnaire. TAU = treatment as usual.*

**Table 1:** Study characteristics of the `r length(table(data_2023_12_4_LSR2_Awks2$studyN))` studies included in the meta-analysis.

Study characteristics of the `r length(table(LSR2weeksStata2insufdata$studyN))` studies not included in the meta-analysis are summarized in Table 2. From these `r length(table(LSR2weeksStata2insufdata$studyN))` studies, there were `r length(table(LSR2weeksStata2insufdata$study))` eligible comparisons.One study reported two comparisons (Greene & Petruzzello, 2022).

```{r echo=FALSE}
knitr::kable(LSR2_studychartab_insuf_140124[, 1:10], "pipe")
```

*Aerobic exercise = physical performance behaviour pattern that increases heart rate and respiration while using large muscle groups repetitively and rhythmically; anaerobic exercise = physical performance behaviour pattern that is performed in short intense bursts with limited oxygen intake; mixed exercise = combination of aerobic and anaerobic exercise.*

**Table 2:** Study characteristics of the `r length(table(LSR2weeksStata2insufdata$studyN))` studies not included in the meta-analysis.

The specific intervention and comparison groups for the `r length(table(data_2023_12_4_LSR2_Awks2$studyN))+length(table(LSR2weeksStata2insufdata$studyN))` eligible studies (`r length(table(LSR2_interventions_150224$study))` comparisons) are presented in table 3.

```{r echo=FALSE}
knitr::kable(LSR2_interventions_150224[, 1:7], "pipe")
```

**Table 3:** Specific interventions for all the included studies

## Risk of bias assessment

The results of the risk of bias (ROB) assessment per domain and study for the primary outcome, PTSD symptom severity, is presented below in Figure 2. Nine studies reported PTSD outcome data post-intervention. Two studies did not report PTSD outcome data (Crombie et al., 2021a; Greene & Petruzzello, 2022). Overall, five studies were rated as high, three studies as some concerns, and one was rated as low risk of bias. High risk of bias was mainly due to deviations from intended intervention (D2)(Voorendonk et al., 2023; Whitworth et al., 2019a; Whitworth et al., 2019b), missing outcome data (D3)(Rosenbaum et al., 2015; Voorendonk et al., 2023; Whitworth et al., 2019a; Whitworth et al., 2019b), and selection of reported results (D5)(Voorendonk et al., 2023; Young-McCaughan et al., 2022). 

```{r echo=F, warning=F, fig.height=8}

rob_traffic_light(data = rob , tool = "ROB2")

```

**Figure 2** Risk of bias for PTSD symptom severity

## Research question I

### Meta-analysis of PTSD Symptom Severity (Primary Outcome)

For the studies included in the meta-analyses, the earliest study was performed in `r format(min((data_2023_12_4_LSR2_Awks2$year)), format="%Y")`, while the most recent study was performed in `r format(max((data_2023_12_4_LSR2_Awks2$year)), format="%Y")`. The median sample size across the studies was `r round(median(data_2023_12_4_LSR2_Awks2$totNrandom, na.rm=T))` participants per study. The median of the mean participant age was `r round(median((data_2023_12_4_LSR2_Awks2$AGEm), na.rm=T))` years (ranging from `r round(min((data_2023_12_4_LSR2_Awks2$AGEm),na.rm=T))` to `r round(max((data_2023_12_4_LSR2_Awks2$AGEm),na.rm=T))` years).

#### Post-intervention (weeks)

`r length(table(data_2023_12_4_LSR2_Awks2$studyN))` studies provided data for PTSD symptom severity and contributed `r length(data_2023_12_4_LSR2_Awks2$tg1PTSD3PTm)` effect measures to the PTSD symptom severity meta-analysis. The summary for PTSD symptom severity is shown in Figure 3.

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=4}
#SUMMARIZE the SMDs from studies using metacont & POOL the data 
LSR2mainMA = metacont(tg1PTSD3PTn , tg1PTSD3PTm ,tg1PTSD3PTsd ,tg2PTSD3PTn, tg2PTSD3PTm, tg2PTSD3PTsd ,
                      data = data_2023_12_4_LSR2_Awks2, studlab = study, sm = "SMD")

forest(LSR2mainMA, 
       sortvar = EXEaug,
       common = FALSE,
       print.I2.ci = T, 
       prediction=T,
       label.left = 'Favours intervention',
       label.right = "Favours control",
       leftcols = c("study", "Intervention", "Comparison"),
       rightcols = c("effect", "ci"),
       fs.study = 10, ff.study = "italic",
       col.diamond = "blue", col.diamond.lines = "black",
       col.square = "turquoise", col.square.lines = "black")
```

**Figure 3:** Meta-analysis of the effects of exercise on PTSD symptom severity.

The meta-analysis found no significant difference in PTSD symptom severity between exercise and comparison groups (`r paste(LSR2mainMA$sm, "=",round(LSR2mainMA$TE.random, 2),", 95% CI:", round(LSR2mainMA$lower.random, 2),",", round(LSR2mainMA$upper.random, 2))`). Low heterogeneity was found as suggested by the prediction interval which is only slightly wider than the confidence interval.

#### Subgroup Analyses and Meta-regression

We explored whether heterogeneity could be potentially explained by differences in study-level characteristics using subgroup analyses and meta-regression below. The following characteristics were explored as sources of heterogeneity for the primary outcome, PTSD symptom severity:

-   Exercise intensity (moderate intensity or high intensity)

-   Specific exercise type (aerobic or anaerobic or mixed)

-   Exercise augmented by treatment as usual (TAU)/therapy (exercise only or exercise + therapy/TAU)

-   Intervention length (weeks)

##### Subgroup analysis by exercise intensity
There was no significant difference in PTSD symptom severity between studies with moderate intensity and those with high intensity exercise groups (see Figure 4). However, visual inspection of the forest plot suggests the effect is larger in the high intensity exercise subgroup.

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=6}
subgroup.intensity = update(LSR2mainMA, subgroup = intensityEX2) 

# produce a forest plot
forest(subgroup.intensity, 
       sortvar = EXEaug,
       common = FALSE,
       print.I2.ci = T, 
       prediction=T,
       label.left = 'Favours intervention',
       label.right = "Favours control",
       leftcols = c("study", "Intervention", "Comparison"),
       rightcols = c( "effect", "ci"),
       fs.study = 10, ff.study = "italic",
       col.diamond = "blue", col.diamond.lines = "black",
       col.square = "turquoise", col.square.lines = "black")
```

**Figure 4:** Sub-group analysis of the effects of exercise on PTSD symptom severity by exercise intensity



##### Subgroup analysis by specific exercise type
The test for subgroup differences found no significant difference in PTSD symptom severity between studies with aerobic or anaerobic or mixed exercise groups (see Figure 5).

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=6}
# create a new R object subgroup.tg1specEX  
subgroup.tg1specEX = update(LSR2mainMA, subgroup = tg1specEX2) 

# produce a forest plot 
forest(subgroup.tg1specEX, 
       sortvar = EXEaug,
       common = FALSE,
       print.I2.ci = T, 
       prediction=T,
       label.left = 'Favours intervention',
       label.right = "Favours control",
       leftcols = c("study", "Intervention", "Comparison"),
       rightcols = c( "effect", "ci"),
       fs.study = 10, ff.study = "italic",
       col.diamond = "blue", col.diamond.lines = "black",
       col.square = "turquoise", col.square.lines = "black")
```

**Figure 5:** Sub-group analysis of the effects of exercise on PTSD symptom severity by specific exercise type

##### Subgroup analysis by exercise alone or tau/therapy augmented by exercise.
There was no significant difference in PTSD symptom severity between studies with exercise alone or TAU/therapy augmented by exercise (see Figure 6). However, visual inspection of the forest plot suggests the effect is larger for exercise alone compared with an inactive comparison group than for the studies that augmented psychotherapy and TAU with exercise.

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=6}
# create a new R object subgroup.EXEaug  
subgroup.EXEaug = update(LSR2mainMA, subgroup = EXEaug2) 

# produce a forest plot using the JAMA layout.
forest(subgroup.EXEaug, 
       sortvar = EXEaug,
       common = FALSE,
       print.I2.ci = T, 
       prediction=T,
       label.left = 'Favours intervention',
       label.right = "Favours control",
       leftcols = c("study", "Intervention", "Comparison"),
       rightcols = c( "effect", "ci"),
       fs.study = 10, ff.study = "italic",
       col.diamond = "blue", col.diamond.lines = "black",
       col.square = "turquoise", col.square.lines = "black")
```

**Figure 6:** Sub-group analysis of the effects of exercise on PTSD symptom severity by exercise alone or TAU/therapy augmented by exercise

##### Meta-regression by intervention length

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=4}

# Do a meta-regression with the continuous covariate intervLENGTH (lenth of intervention - weeks) 

 intervLENGTHreg = metareg(LSR2mainMA, intervLENGTH) 

```

Overall, there is no strong evidence that the intervention length affected the treatment effect. The meta-regression analysis yielded a coefficient of `r round(intervLENGTHreg$beta[2], 2)` (95% CI: `r round(intervLENGTHreg$ci.lb[2], 2)`, `r round(intervLENGTHreg$ci.ub[2], 2)`).

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=4}

# generate a bubble plot. 

bubble(intervLENGTHreg, col.line = "blue") 

```

**Figure 7:** Meta-regression of the effects of exercise on PTSD symptom severity by intervention length

### Meta-analysis of Treatment Dropout (secondary outcome)

#### Post-intervention (weeks)

`r length(table(data_2023_12_4_LSR2_Awks2$studyN))` studies provided data for treatment dropout, and contributed `r length(data_2023_12_4_LSR2_Awks2$tg1Ndrop)` effect measures to the treatment dropout meta-analysis. The forest plot for the risk of treatment dropout is shown in Figure 8.

```{r echo=F, fig.fullwidth=T, fig.width=11, fig.height=4}
#Summarize the study-specific risk ratios for study completors from studies using metabin
#create an object of class meta called “pooledRRdrop”
pooledRRdrop = metabin(tg1Ndrop, tg1Nrandom,	tg2Ndrop,tg2Nrandom,
                       data = data_2023_12_4_LSR2_Awks2, studlab = study, sm = "RR")

#obtain the forest plot 
forest(pooledRRdrop, 
       sortvar = EXEaug,
       common = FALSE,
       print.I2.ci = T, 
       prediction=T,
       label.left = 'Favours intervention',
       label.right = "Favours control",
       leftcols = c("study", "Intervention", "Comparison"),
       rightcols = c("effect", "ci"),
       fs.study = 10, ff.study = "italic",
       col.diamond = "blue", col.diamond.lines = "black",
       col.square = "turquoise", col.square.lines = "black")


```

**Figure 8:** Meta-analysis of the dropout rates between the intervention and control groups

***
***This is not matching the forrest plot***

There is no evidence of a difference in treatment dropout between exercise and comparison groups (`r paste(pooledRRdrop$sm, "=" ,round(pooledRRdrop$TE.random, 2), ", 95% CI: ", round(pooledRRdrop$lower.random, 2), ",", round(pooledRRdrop$upper.random, 2))`) and there is large heterogeneity, as shown by the prediction interval (`r paste(round(exp(pooledRRdrop$lower.predict),2), "to",round(exp(pooledRRdrop$upper.predict),2))`).

***

### Sensitivity Analyses
```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=4}
#SUMMARIZE the SMDs from studies using metacont 
rob.sens.MA = metacont(tg1PTSD3PTn , tg1PTSD3PTm ,tg1PTSD3PTsd ,tg2PTSD3PTn, tg2PTSD3PTm, tg2PTSD3PTsd ,
                      data = data.rob.sens, studlab = study, sm = "SMD")
```
We examine the robustness of the findings for the primary outcome by excluding studies with high risk of bias. `r length(table(data.rob.sens$study))` studies included in the meta-analyses were rated as low or some concerns. When restricting the analysis to studies with moderate or low risk of bias, the effect of exercise on PTSD symptoms severity is `r paste(rob.sens.MA$sm, "=", round(rob.sens.MA$TE.random, 2),"(95% CI:", round(rob.sens.MA$lower.random, 2), ",", round(rob.sens.MA$upper.random, 2))`). For reference, the main effect size for the primary outcome is `r paste(LSR2mainMA$sm, "=", round(LSR2mainMA$TE.random, 2),"(95% CI:", round(LSR2mainMA$lower.random, 2), ",", round(LSR2mainMA$upper.random, 2))`), so the results do not change substantially.

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=4}
#POOL the data 
forest(rob.sens.MA, 
       sortvar = EXEaug,
       common = FALSE,
       print.I2.ci = T,
       label.left = 'Favours intervention',
       label.right = "Favours control",
       leftcols = c("study", "Intervention", "Comparison"),
       rightcols = c("effect", "ci"),
       fs.study = 10, ff.study = "italic",
       col.diamond = "blue", col.diamond.lines = "black",
       col.square = "turquoise", col.square.lines = "black")
```

**Figure 9:** Meta-analysis of the effects of exercise on PTSD symptom severity when excluding studies with high risk of bias

### Reporting bias

The forest plot below shows the meta-analysis results of the primary outcome ordered by the precision of the studies. It seems that the smaller studies showed larger effects favouring the intervention.

```{r echo=FALSE, fig.fullwidth=T, fig.width=11, fig.height=4}
# forest polot ordered by precision

forest(LSR2mainMA, 
       sortvar = seTE,
       common = FALSE,
       print.I2.ci = T, 
       prediction=T,
       label.left = 'Favours intervention',
       label.right = "Favours control",
       leftcols = c("study", "Intervention", "Comparison"),
       rightcols = c("effect", "ci"),
       fs.study = 10, ff.study = "italic",
       col.diamond = "blue", col.diamond.lines = "black",
       col.square = "turquoise", col.square.lines = "black")
```

**Figure 10:** Forest plot of the meta-analysis results of the primary outcome ordered by the precision of the studies

## Summary of the evidence on the primary outcome from the human studies

The primary outcome was efficacy in reducing overall PTSD symptom severity in patients with PTSD. The summary of the evidence on PTSD symptom severity outcome for PTSD is reported below.

|                                           |                                                   |                                                                                                                                                                                                                                              |                                        |                                  |                                                          |
|------------|------------|------------|------------|------------|------------|
| **Source of evidence**                    | **Summary of the association**                    | **Bias due to study limitations**                                                                                                                                                                                                            | **Bias due to reporting bias**         | **Bias due to indirectedness**   | **Bias due to other reasons**                            |
| Exercise intervention vs comparison group | n = 9 , k = 700 , SMD = -0.08, 95%CI: -0.24, 0.07 | Overall, five studies were rated as high, three studies as some concerns, and one was rated as low risk of bias. High risk of bias due to to deviations from intended intervention, missing outcome data, and selection of reported results. | Two studies rated as high risk of bias | *what here*                      | Lack of blinding of participants and outcomes assessors. |
